Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2017-06-03 chap-intro.tex
Intro: Clarify parts and responsibilities

In particular, this thesis is divided into two parts:
\begin{itemize}
\item we first describe, in \cref{sec:ch-aosd13}, work on optimizing collection queries by static program transformation.
\item in the remaining chapters, we describe work on incrementalizing programs
  by static program transformation. This thesis presents the first approach that
  handles higher-order programs by using program transformation; hence, while
  our main examples use collection queries, we phrase the work in terms of
  $\lambda$-calculi with unspecified primitives.
\end{itemize}

\pg{rewrite this once we have a second part.}
The second part makes the following contributions:
\begin{itemize}
\item We present a novel mathematical theory of changes and derivatives, which is more
  general than other work in the field because changes are
  first-class entities, they are distinct from base values and
  they are defined also for functions.\pg{readd some reference.}
  %(\cref{sec:1st-order-changes}).
  %KO: I think the next sentence cannot be understood at this point.
  %We introduce changes for complex types, defined compositionally.
%
\item We present the first approach to incremental computation for pure
  $\lambda$-calculi by a source-to-source transformation, $\DERIVE$, that
  requires no run-time support. The transformation produces an incremental
  program in the same language; all optimization techniques for the original
  program are applicable to the incremental program as well.
%KO: commented this out. I think the purity is not important enough
%to deserve another sentence here, since we only vaguely hint
%at "further research".
%Since our incremental programs use no impure features, they are
%especially amenable to further optimizations, making this approach
%very suitable for further research.
%
% KO: Let's have one bullet point per section. Also, a conjecture
% sounds like a rather weak contribution
%\item We argue that incrementalization is efficient on
%  \emph{self-maintainable programs}, and discuss how further research on
%  static or dynamic memoization can speed up a larger class of programs (\cref{sec:performance-cons}).
%  \pg{This contribution references text which is now commented
%    out. I believe the text should be brought back in.}
%
We prove that our incrementalizing transformation $\DERIVE$
is correct~(\cref{eq:correctness})
by a machine-checked formalization in Agda~\citep{agda-head}.
The proof gives insight into the definition of $\DERIVE$: we
first construct the derivative $\EvalInc{-}$ of the denotational
semantics of a simply-typed $\lambda$-calculus term, that is, its
\emph{change semantics}.
%
Then, we show that $\DERIVE$ is produced by erasing
$\EvalInc{-}$ to a simply-typed program.% (\cref{sec:correctness}).

\item While we focus mainly on the theory of changes
and derivatives, we also perform a performance case study.
We implement the derivation transformation in Scala,
with a plug-in architecture that can be extended with new base
types and primitives. We define a plugin with support for
different collection types and use the plugin to
incrementalize a variant of the MapReduce programming model~\citep{Lammel07}.
  Benchmarks show that on this program,
  incrementalization can reduce asymptotic complexity and can turn $O(n)$
  performance into $O(1)$, improving running time by over 4
  orders of magnitude on realistic inputs (\cref{sec:applying}).

\end{itemize}

\pg{Readdd section on excluded papers.
% \section{Excluded papers}
% During my PhD work I collaborated on several other papers.

@@

% \paragraph{DSL work}
% While working on modular embedded DSLs I collaborated to \citet{ErdwegGR12},
% \pg{TODO GADT paper}, \pg{TODO maybe manuscript on cake pattern?}.
}

>>>>

In particular, this thesis is divided into two parts:
\begin{itemize}
\item in \cref{part:ch-aosd13}, we describe work on optimizing collection queries by static program transformation~\citep{GiarrussoAOSD13}.
\item in \cref{part:incr}, we describe work on incrementalizing programs
  by static program transformation. This thesis presents the first approach that
  handles higher-order programs by using program transformation; hence, while
  our main examples use collection queries, we phrase the work in terms of
  $\lambda$-calculi with unspecified primitives~\citep{CaiEtAl2014ILC}.
\end{itemize}

\pg{rewrite this once we have a second part.}
The second part makes the following contributions:
\begin{itemize}
\item We present a novel mathematical theory of changes and derivatives, which is more
  general than other work in the field because changes are
  first-class entities, they are distinct from base values and
  they are defined also for functions.\pg{readd some reference.}
  %(\cref{sec:1st-order-changes}).
  %KO: I think the next sentence cannot be understood at this point.
  %We introduce changes for complex types, defined compositionally.
%
\item We present the first approach to incremental computation for pure
  $\lambda$-calculi by a source-to-source transformation, $\DERIVE$, that
  requires no run-time support. The transformation produces an incremental
  program in the same language; all optimization techniques for the original
  program are applicable to the incremental program as well.
%KO: commented this out. I think the purity is not important enough
%to deserve another sentence here, since we only vaguely hint
%at "further research".
%Since our incremental programs use no impure features, they are
%especially amenable to further optimizations, making this approach
%very suitable for further research.
%
% KO: Let's have one bullet point per section. Also, a conjecture
% sounds like a rather weak contribution
%\item We argue that incrementalization is efficient on
%  \emph{self-maintainable programs}, and discuss how further research on
%  static or dynamic memoization can speed up a larger class of programs (\cref{sec:performance-cons}).
%  \pg{This contribution references text which is now commented
%    out. I believe the text should be brought back in.}
%
\item We prove that our incrementalizing transformation $\DERIVE$
is correct~(\cref{eq:correctness}) by a novel machine-checked logical relation
proof, mechanized in Agda~\citep{agda-head}.

\item While we focus mainly on the theory of changes
and derivatives, we also perform a performance case study.
We implement the derivation transformation in Scala,
with a plug-in architecture that can be extended with new base
types and primitives. We define a plugin with support for
different collection types and use the plugin to
incrementalize a variant of the MapReduce programming model~\citep{Lammel07}.
  Benchmarks show that on this program,
  incrementalization can reduce asymptotic complexity and can turn $O(n)$
  performance into $O(1)$, improving running time by over 4
  orders of magnitude on realistic inputs (\cref{sec:applying}).
\end{itemize}

\subsection{Included papers}
This thesis includes material from joint work with colleagues.

\Cref{part:ch-aosd13} is based on work by \citet*{GiarrussoAOSD13}. While the
work was in collaboration, a few clearer responsibilities arose.
I did most of the implementation work: among other things I devised the
embedding for collection operations, implemented optimizations and indexing,
implemented interpretation when compilation did not work well, and performed the
performance evaluation.
% Michael Eichberg and Ralf Mitschke contributed to the
% evaluation on FindBugs code.
% Christian K{\"{a}}stner contributed, among other things,
% to the experiment design for the performance evaluation.
% Klaus Ostermann proposed the original idea and supervised the project.

\Cref{part:incr} is originally based on work by \citet*{CaiEtAl2014ILC}, though
significantly revised. This work was even more of a team effort. I designed the
overall project and came up with the original notion of change structures; Cai
Yufei contributed differentiation itself and its first correctness proof.
\Cref{part:incr} contains a novel correctness proof; its history and
contributions are discussed in \cref{sec:ilc-dev-history}.

Furthermore, \cref{part:static} comes from a new, unpublished manuscript. I
designed the overall approach, the transformation and the first case study on
sequences, defining the change structure, designing the primitives and the first
case study. Yann Régis-Gianas designed the proofs. Philipp Schuster contributed
to the evaluation.

\pg{Readd section on excluded papers.}
% \subsection{Excluded papers}
% During my PhD work I collaborated on several other papers.

@@

% \paragraph{DSL work}
% While working on modular embedded DSLs I collaborated to \citet{ErdwegGR12},
% \pg{TODO GADT paper}, \pg{TODO maybe manuscript on cake pattern?}.
